On Sample Size for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2023-05-16 Manuscript/manuscript.tex
5/15 Manuscript Update

The goal of this paper is to provide some practical recommendations on
necessary sample size for block bootstrap with dependent data. We consider a
simple situation of a stationary time series, where the parameters of
interests are the mean, standard deviation, and the first-order
autocorrelation coefficient. We compare five variants of block bootstrap
CIs from the literature \citep{diciccio1996bootstrap,
  rice2006mathematical}: a basic percentile interval, a percentile
interval centered around the estimate, a bias-corrected and
accelerated ($BC_a$) interval, a standard normal interval, and a student's $t$
interval. Their
empirical coverage rates at different sample sizes and dependence levels are
compared in a simulation study. The results of this study suggest that
appropriate recovery of temporal dependence parameters is heavily dependent on
the type of interval used.

@@

The block size~$l$ needs to be chosen with care. It should be large enough for
each bootstrap sample to preserve the serial dependence, yet small enough for
there to be a large number of blocks to give sufficient variability between
each bootstrap sample. As $n$ increases, both~$l$
and $n / l$ should also increase. To achieve this, $l$ is
often assigned a value as a function of $n$. A common function that is
considered optimal is $l = \lceil n^{1/3} \rceil$
\citep{buhlmann1999block}, which was adopted in this
study. It may be worth investigating in a future study if another block size
scheme is \jy{Seems out of place. Maybe in the discussion section.}
optimal for the context of a time series by comparing different block sizes
whilst keeping the sample size fixed.

@@

\jy{order them as we discussed.}

@@

\paragraph{Standard Normal CI}

A $(1 - \alpha$ standard normal CI depends on the standard error $\hat{se}$
for each of the
$B$ bootstrap point estimates
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$. 
\[
(\hat{\theta}_{n} - z_{(1-\alpha/2)}*\hat{se}, \hat{\theta}_{n} -
z_{(\alpha/2)}*\hat{se}).
\]
This CI is centered by the point estimator $\hat\theta_n$ and is symmetric.
The standard
normal CI is a table-based bootstrap interval approach, which is to say that
it relies on the 
assumption that $\hat\theta_n$ is normally distributed and the expected value
of $\hat\theta_n$
is $\theta$.

\paragraph{Student's $t$ CI}
A $(1 - \alpha$ student's $t$ CI is
\[
(\hat{\theta}_{n} - t_{(1-\alpha/2), k - 1}*\hat{se}, \hat{\theta}_{n} -
t_{(\alpha/2), k -1}*\hat{se}),
\]
where k is the number of blocks.
This CI is centered by the point estimator $\hat\theta_n$ and is symmetric.
Like the standard 
normal interval, the student's t CI is a table-based bootstrap interval
approach. In this case,
we assume that $\hat\theta_n$ follows the $t_{k-1}$ distribution with an
expected value of $\theta$.

@@

Three target parameters of $X_t$ were considered:
1) $\mu = 0$, the mean of $X_t$;
2) $\sigma_x = 1$, the standard deviation of $X_t$; and
2) $\phi$, the lag-1 auto-correlation coefficient.
To investigate the effect of sample size~$n$, we considered an array of values
$n \in \{100, 200, 400, 800, 1600\}$. In each configuration, we generated 10000
replicates. For each replicate, we constructed three 95\% block bootstrap
CIs for each parameter as described in the last section. We can estimate 
$\mu$, $\sigma_x$, and $\phi$ by computing the sample mean, sample 
standard deviation,
and lag-1 auto-correlation, respectively, of each bootstrap sample, and then 
constructing intervals for each parameter using the appropriate procedures
described in Section~\ref{sec:bbci} .
Then we estimated their actual coverage rates along with their 95\% confidence
intervals from the 10000 replicates. The block bootstrap sampling step was done
with function \texttt{tsboot} from R package \textsl{boot} \citep{boot}, with
block size $\lceil n / l \rceil$. This function by default is an implementation
of moving block bootstrap as described in the previous section, meaning that
that blocks are allowed to wrap around, and $l = \lceil n^{1/3} \rceil$.
\jy{size? also it was overlapping blocks. Describe the details here.}
\mc{addressed}
\eds{Is $l = \lceil n^{1/3} \rceil$? Is it also worth investigating different 
choices of $l$ for fixed (large) $n$?}
\mc{Addressed in the Block Bootstrap Sample subsection}
\eds{how did you estimate $\mu$, $\sigma_x$ and $\phi$?}
\mc{addressed}

@@

Based on Figure~\ref{fig:mu}, 
the performance is about the same for Percentile, Centered, BCA, and 
Standard CIs with respect to estimating $\mu$. They more or less need
about the same sample size to have the correct coverage. For $\phi = -0.4$, 
they slightly over-cover $\mu$ for all sample sizes, but the error bars generally 
hover around the $95\%$ line. For $\phi = -0.2$, all four cover $\mu$ at the nominal 
rate when $n = 400$, and when $n = 1600$, although there seems to be a dip in 
coverage for $n = 800$. However, this deviation in trend can most likely be 
attributed to noise. Besides that, the coverage rates remain above $.9375$ 
for all sample sizes for this level of $\phi$. For $\phi = 0$, or when the 
sample is independently normally distributed, $n = 1600$ seems to 
be enough to recover $\mu$. For $\phi = 0.2$, the coverage rates are above 
$.9375$ when $n = 800$ and $n = 1600$, but the coverage intervals do not contain
$95\%$. Interestingly, there seems to be a drop in coverage rate from 
$n = 800$ to $n = 1600$. For $\phi = 0.4$, the coverage rates are above $.925$
when $n = 800$ and $n = 1600$, and there is a more noticeable downward 
trend from $n = 800$ to $n = 1600$. Student's $t$ CI has noticeably higher coverage
than the other CIs. For $\phi = -0.4$, the Student's $t$ CI actually 
over-covers $\mu$ for all sample sizes, and the lower error bars
are above the $95\%$ line for all sample sizes. For $\phi = -0.2$, it still slightly 
over-covers $\mu$ for all sample sizers, but the error bars generally hover
lower and around the $95\%$ line. For $\phi = 0$, the interval for the 
coverage rate contains $.95$ for all sample sizes. For $\phi = 0.2$ and 
$\phi = 0.4$, the trend in coverage is similar to the other CIs, but 
the overall coverage rates are higher, although they still do not reach the $95\%$
line.

Based on Figure~\ref{fig:sigma}, 
$BC_a$ CIs greatly undercover $\sigma_x$ for all sample sizes, although
there is no trend of coverage deterioration, indicating that there may eventually
a sample size large enough for $\sigma_x$ to be covered at the nominal level.
However, because it's performance is so much poorer than the other CIs,
$BC_a$ should not be used. Percentile, Centered, and Standard CIs perform similarly,
with respect to estimating $\sigma_x$. For $\phi = -0.4$, $\phi = -0.2$, and $\phi = 0$, 
$n = 1600$ seems to be large enough to cover $\sigma_x$ at the nominal level
when using the Standard Normal CI interval, while the others are very close, hovering
around $.945$ for $n = 1600$. For $\phi > 0$, a larger sample size does seem
to be necessary to cover $\sigma_x$ at the nominal rate. However, for $\phi = 0.2$, 
when $n \geq 400$, coverage is above $.93$ for all three of these CIs. For $\phi = 0.4$,
when $n = 1600$, coverage is above $.93$ for all three of these CIs.
Like for $\mu$, Student's $t$ CIs have the highest overall coverage of $\sigma_x$.
Additionally, the Student's $t$ method does not over-cover $\sigma_x$ for any $n$.
For $\phi = -0.4$, $\sigma_x$ is covered at the nominal rate for $n \geq 1600$.
For $\phi = -0.2$ and $\phi = 0$, $\sigma_x$ is covered at the nominal rate for $n \geq 400$.
For $\phi = 0.2$, although the coverage does not reach the $95\%$ line by $n = 1600$, 
it seems to be around $.945$ for $n \geq 400$
For $\phi = 0.4$, although the coverage does not reach the $95\%$ line by $n = 1600$, 
it seems to be above $.93$ for $n \geq 400$

Based on Figure~\ref{fig:phi}, 
Percentile and $BC_a$ CIs deteriorate clearly for any level of temporal correlation,
so they should not be used. The deterioration becomes more evident
as the absolute value of $\phi$ increases, regardless of whether $\phi$ is 
negative or positive. Centered and Standard CIs perform similarly;
a sample size of $n = 800$ appears to be sufficient to recover $\phi$ at the 
nominal level when using either of these two CIs. 
Student's $t$ CIs appear to recover $\phi$ at the nominal level for $n \geq 100$ for all 
$\phi$.
\jy{One summary paragraph at the end.}
\mc{paragraph below}

To summarize, the CI type chosen should be based on the target parameter. 
When estimating $\mu$, the Student's t CI seems to have the highest coverage
overall, although that leads to over-coverage when $\phi$ is -0.4. For a
sample with no temporal dependence, the Student's t CI cover $\mu$ perfectly
at the nominal rate even for small sample sizes. For positive $\phi$, a larger
sample size appears to be necessary regardless of the CI used. In the case
that $\phi$ is unknown and expected to be greater than $0$, $n > 1600$ may be
necessary for block bootstrap interval estimation of $\mu$. If that is not the
case, a lesser sample size is likely sufficient. When estimating $\sigma$,
$BC_a$ should not be used, and Student's t again has the greatest coverage
rates. In the case that $\phi$ is unknown and expected to be greater than $0$,
$n > 1600$ may be necessary for block bootstrap interval estimation of
$\sigma$. If that is not the case, a lesser sample size is likely sufficent.
When estimating $\phi$, Percentile and $BC_a$ CIs 
deteriorate in coverage, indicating they should not be used in this context. 
Increasing n will not fix the performance of either of these intervals 
because the assumption that the coverage rate will approach the nominal level
as n increases is invalid. Centered and Standard CIs managed to fix the
deterioration problem, and they perform similarly, but Student's t covers
$\phi$ at the nominal rate for small sample sizes regardless of the absolute
value or sign of $\phi$. Interestingly, $n = 100$ seems to be large enough to
recover $\phi$ when using the Student's t method.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_mu}
  \caption{Plots of $\mu$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600$ were recorded.}
  \label{fig:mu}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_sigma}
  \caption{Plots of $\sigma_x$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600$ were recorded.}
  \label{fig:sigma}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_phi}
  \caption{Plots of $\phi$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600$ were recorded.}
  \label{fig:phi}
\end{figure}

@@

When using $\hat{\theta}_{n}$-centered intervals and $\phi$ is
unknown, the results of this study suggest that an $n$ of around $1600$ may be
necessary for common practice to estimate a simple parameters like $\mu$ 
or $\sigma_x$. If $\phi$ is already known, a lesser $n$ may be adequate
to
estimate these parameters. ... However, in real world applications, $\phi$ is
more commonly found to be positive, so a larger $n$ is more likely to be
necessary. Lastly, to estimate $\phi$, an $n$ of around $100$ using the Student's $t$ method
 may be sufficient
Further investigation may be necessary to see if there are other interval
corrections that fix the coverage deterioration problem for $\phi$.

>>>>

The goal of this paper is to provide some practical recommendations on
necessary sample size for block bootstrap with dependent data. We consider a
simple situation of a stationary time series, where the parameters of
interests are the mean, standard deviation, and the first-order
autocorrelation coefficient. We compare five variants of block bootstrap
CIs from the literature \citep{diciccio1996bootstrap,
  rice2006mathematical}: a basic percentile interval, a percentile
interval centered around the estimate, a bias-corrected and
accelerated ($BC_a$) interval, a standard normal interval, and a Student's $t$
interval. Their
empirical coverage rates at different sample sizes and dependence levels are
compared in a simulation study. The results of this study suggest that
appropriate recovery of temporal dependence parameters is heavily dependent on
the type of interval used.

@@

The block size~$l$ needs to be chosen with care. It should be large enough for
each bootstrap sample to preserve the serial dependence, yet small enough for
there to be a large number of blocks to give sufficient variability between
each bootstrap sample. As $n$ increases, both~$l$
and $n / l$ should also increase. To achieve this, $l$ is
often assigned a value as a function of $n$. A common function that is
considered optimal is $l = \lceil n^{1/3} \rceil$
\citep{buhlmann1999block}, which was adopted in this
study.

@@

\jy{order them as we discussed.}
\mc{addressed}
\paragraph{Standard Normal CI}
The procedure for constructing a Standard Normal CI is described in 
\citet{efron1993introduction}.
A $(1 - \alpha$ standard normal CI depends on the standard error $\hat{SE}$
for each of the
$B$ bootstrap point estimates
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$. 
\[
(\hat{\theta}_{n} - z_{(1-\alpha/2)}\hat{SE}, \hat{\theta}_{n} -
z_{(\alpha/2)}\hat{SE}).
\]
This CI is centered by the point estimator $\hat\theta_n$ and is symmetric.
The standard
normal CI is a table-based bootstrap interval approach, which is to say that
it relies on the 
assumption that $\hat\theta_n$ is normally distributed and the expected value
of $\hat\theta_n$
is $\theta$.

\paragraph{Student's $t$ CI}
The procedure for constructing a Student's $t$ CI is described in 
\citet{efron1993introduction}.
A $(1 - \alpha$ Student's $t$ CI is
\[
(\hat{\theta}_{n} - t_{(1-\alpha/2), k - 1}\hat{SE}, \hat{\theta}_{n} -
t_{(\alpha/2), k -1}\hat{SE}),
\]
where $k$ is the number of blocks.
This CI is centered by the point estimator $\hat\theta_n$ and is symmetric.
Like the standard 
normal interval, the Student's $t$ CI is a table-based bootstrap interval
approach. In this case,
we assume that $\hat\theta_n$ follows the $t_{k-1}$ distribution with an
expected value of $\theta$.

@@

@@

Three target parameters of $X_t$ were considered:
1) $\mu = 0$, the mean of $X_t$;
2) $\sigma_x = 1$, the standard deviation of $X_t$; and
2) $\phi$, the lag-1 auto-correlation coefficient.
To investigate the effect of sample size~$n$, we considered an array of values
$n \in \{100, 200, 400, 800, 1600, 3200\}$. In each configuration, we generated 10000
replicates. For each replicate, we constructed three 95\% block bootstrap
CIs for each parameter as described in the last section. We can estimate 
$\mu$, $\sigma_x$, and $\phi$ by computing the sample mean, sample 
standard deviation,
and lag-1 auto-correlation, respectively, of each bootstrap sample, and then 
constructing intervals for each parameter using the appropriate procedures
described in Section~\ref{sec:bbci} .
Then we estimated their actual coverage rates along with their 95\% confidence
intervals from the 10000 replicates. The block bootstrap sampling step was done
with function \texttt{tsboot} from R package \textsl{boot} \citep{boot}, with
block size $\lceil n / l \rceil$. This function by default is an implementation
of moving block bootstrap as described in the previous section, meaning that
that blocks are allowed to wrap around, and $l = \lceil n^{1/3} \rceil$.
\jy{size? also it was overlapping blocks. Describe the details here.}
\mc{addressed}
\eds{Is $l = \lceil n^{1/3} \rceil$? Is it also worth investigating different 
choices of $l$ for fixed (large) $n$?}
\mc{Addressed in the Block Bootstrap Sample subsection}
\eds{how did you estimate $\mu$, $\sigma_x$ and $\phi$?}
\mc{addressed}

@@

Based on Figure~\ref{fig:mu}, 
Student's $t$ CIs appear to have the greatest coverage rates with respect 
to estimating $\mu$. In fact, for $\phi = -0.4$, Student's $t$ CIs over-cover $\mu$
for all sample sizes, whereas other CIs have coverage rates that
hover around the $95\%$ line at this level of $\phi$. For $\phi = -0.2$, 
Standard, Percentile, Centered, and $BC_a$ CIs cover $\mu$ at the nominal 
rate when $n \geq 400$. Student's $t$ CIs either recover or over-cover $\mu$
for $n \geq 100$. For $\phi = 0$, or when the 
sample is independently normally distributed, $n = 800$ seems to 
be enough to recover $\mu$ when using Standard, Percentile, Centered, or $BC_a$ 
CIs. Student's $t$ CIs cover $\mu$ at the nominal rate for $n \geq 100$
For $\phi = 0.2$, $n = 3200$ seems to be large enough when using
Standard, Student's $t$, or Centered CIs, but Percentile and $BC_a$ CIs undercover
$\mu$ for all sample sizes at this level of $\phi$.
For $\phi = 0.4$, the Student's $t$ is the only type of CI that is able to recover $\mu$
for a sample as large as $n = 3200$. The other CIs undercover $\mu$ for all sample 
sizes at this level of $\phi$. Unsurprisingly, a greater level of $\phi$ demands a 
larger $n$ to recover $\mu$ at the nominal rate.

Based on Figure~\ref{fig:sigma}, 
$BC_a$ CIs greatly undercover $\sigma_x$ for all sample sizes, although
there is no trend of coverage deterioration, indicating that there may eventually
be a sample size large enough for $\sigma_x$ to be covered at the nominal level.
For $\phi = -0.4$, Standard Normal and Student's $t$ CIs recover $\sigma$ when 
$n = 3200$, whereas as the other CIs undercover $\sigma$ for all sample sizes.
For $\phi  = -0.2$, all CIs except for $BC_a$ are able to cover $\sigma$ at the 
nominal rate for $n \geq 1600$, and Student's $t$ CIs cover $\sigma$ at the 
nominal rate when $n \geq 800$. When $\phi = 0$, $n \geq 1600$ seems to 
be large enough if Standard and Student's $t$ CIs are used. For $\phi = 0.2$, 
consistent coverage of $\sigma$ at the nominal level appears to only be true
for Student's $t$ CIs for $n \geq 1600$. For $\phi = 0.4$, the Student's $t$ CI is 
able to reach nominal coverage of $\sigma$ when $n = 1600$, although in 
our results, coverage appears to drop below the nominal rate for $n = 3200$, 
indicating that perhaps an even larger $n$ is needed to recover $\sigma$ at
this level of $\phi$.

Based on Figure~\ref{fig:phi}, 
Percentile and $BC_a$ CIs deteriorate clearly for any level of temporal correlation,
so they should not be used. The deterioration becomes more evident
as the absolute value of $\phi$ increases, regardless of whether $\phi$ is 
negative or positive. Centered and Standard CIs perform similarly;
a sample size of $n = 800$ appears to be sufficient to recover $\phi$ at the 
nominal level when using either of these two CIs. 
Student's $t$ CIs appear to recover $\phi$ at the nominal level for $n \geq 100$ for all 
$\phi$.

\jy{One summary paragraph at the end.}
\mc{paragraph below}

To summarize, the CI type chosen should be based on the target parameter. 
When estimating $\mu$, the Student's $t$ CI seems to have the highest coverage
overall, although that leads to over-coverage when $\phi$ is -0.4. For a
sample with no temporal dependence, the Student's $t$ CI cover $\mu$ perfectly
at the nominal rate even for small sample sizes. For positive $\phi$, a larger
sample size appears to be necessary regardless of the CI used. In the case
that $\phi$ is unknown and expected to be greater than $0$, $n = 3200$ may be
necessary for block bootstrap interval estimation of $\mu$. If $\phi$ is known,
a lesser sample size may be sufficient. When estimating $\sigma$,
$BC_a$ should not be used, and Student's $t$ again has the greatest coverage
rates. In the case that $\phi$ is unknown and expected to be greater than $0$,
$n > 3200$ may be necessary for block bootstrap interval estimation of
$\sigma$. Again, if $\phi$ is known, a lesser sample size may be enough.
When estimating $\phi$, Percentile and $BC_a$ CIs 
deteriorate in coverage, indicating they should not be used in this context. 
Increasing $n$ will not fix the performance of either of these intervals 
because the assumption that the coverage rate will approach the nominal level
as $n$ increases is invalid. Centered and Standard CIs managed to fix the
deterioration problem, and they perform similarly, but Student's $t$ covers
$\phi$ at the nominal rate for small sample sizes regardless of the absolute
value or sign of $\phi$. Interestingly, $n = 100$ seems to be large enough to
recover $\phi$ when using the Student's $t$ method.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_mu}
  \caption{Plots of $\mu$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600, 3200$ were recorded.}
  \label{fig:mu}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_sigma}
  \caption{Plots of $\sigma_x$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600, 3200$ were recorded.}
  \label{fig:sigma}
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_phi}
  \caption{Plots of $\phi$ coverage rates for each interval type and $\phi$
    combination. The orange dotted line represents the 95\% nominal coverage
    rate. The error bars represent a 95\% CI for the coverage
    proportion out of 10000 replications for each $n$. Results for 
    $n = 100, 200, 400, 800, 1600, 3200$ were recorded.}
  \label{fig:phi}
\end{figure}

@@

When using $\hat{\theta}_{n}$-centered intervals and $\phi$ is
unknown, the results of this study suggest that an $n$ of around $3200$ may be
necessary for common practice to estimate a simple parameters like $\mu$ 
or $\sigma_x$. If $\phi$ is already known, a lesser $n$ may be adequate
to
estimate these parameters. ... However, in real world applications, $\phi$ is
more commonly found to be positive, so a larger $n$ is more likely to be
necessary. Lastly, to estimate $\phi$, an $n$ of around $100$ using the 
Student's $t$ method
 may be sufficient
Further investigation may be necessary to see if there are other interval
corrections that fix the coverage deterioration problem for $\phi$.
